---
title: "PEH-OUD_Microsimulation"
format: docx
---

```{r setup, include=FALSE}
rm(list = ls())
```

# Model inputs and settings

## Model settings

```{r}
n.i <- 1000         # Number of simulated individuals
n.t <- 1200         # Number of cycles (each cycle = 1 month = 1/12 year)
# n.t <- 600
annual_d.c <- 0.03  # Annual discount rate for costs
annual_d.e <- 0.03  # Annual discount rate for QALYs
d.c <- (1 + annual_d.c) ^ (1/12) - 1
d.e <- (1 + annual_d.e) ^ (1/12) - 1
```

## Definition of states

```{r}
v.housing <- c("H", "S", "U")           # Housing states: Permanently housed, Sheltered, Unsheltered
v.OUD     <- c("NU", "YT", "NT", "AT")  # OUD states: Never users, OUD in treatment, OUD out of treatment, Abstinent
```

## Conversion functions

### Rate to probability

```{r}
rate_to_prob <- function(r, t = 1/12){
  if (any(r < 0)){
    stop("Rate is not greater than or equal to 0.")
  }
  p <- 1 - exp(- r * t)
  return(p)
}
```

### Probability to rate

```{r}
prob_to_rate <- function(p, t = 1){
  if (any(p > 1) || any(p < 0)){
    stop("Probability is not between 0 and 1.")
  }
  r <- -(1/t)*log(1 - p)
  return(r)
}
```

### Probability to probability

```{r}
prob_to_prob <- function(p, t = 1){
  if (any(p > 1) || any(p < 0)){
    stop("Probability is not between 0 and 1.")
  }
  p_new <- 1 - (1-p)^(t)
  return(p_new)
}
```

## Initial population characteristics

```{r}
set.seed(123)

age_groups <- c("18-24", "25-39", "40-49", "50-64", "65+")
init_age   <- sapply(sample(age_groups, n.i, replace = TRUE,
                          prob = c(0.04, 0.29, 0.24, 0.37, 0.06)),
                   function(group) switch(  # Function to sample a specific age uniformly within the chosen group
                     group,
                     "18-24" = sample(18:24, 1),
                     "25-39" = sample(25:39, 1),
                     "40-49" = sample(40:49, 1),
                     "50-64" = sample(50:64, 1),
                     "65+" = sample(65:80, 1)))
init_sex   <- sample(c("M", "F"), n.i, replace = TRUE, prob = c(0.7, 0.3))
p_S <- 22.4 / 100
p_U <- (56.4 + 21.2) / 100
init_housing <- sample(c("S", "U"), n.i, replace = TRUE, prob = c(p_S, p_U))
init_OUD     <- sample(c("NU", "YT", "NT", "AT"), n.i, replace = TRUE, prob = c(0, 0, 1, 0))
init_dead    <- rep(0, n.i)  # 0 = alive; 1 = dead
```

## Mortality table and functions

### CDC life table

```{r}
mortality_df <- read.csv("CDC_Life_Table.csv")
```

### Prevalence

#### OUD prevalence

```{r}
p_curr_reg <- 10.4 / 100
p_curr_occ <- 3.7 / 100
p_life_reg <- 20.4 / 100
p_life_any <- 29.0 / 100

p_NU <- 1 - p_life_reg
p_AT <- 1 - p_NU - p_curr_reg
p_OUD_YT <- (0.047 * p_life_reg) / p_curr_reg  # Assume all in treatment were regular users in past 6 months
p_OUD_NT <- 1 - p_OUD_YT
```

#### Housing prevalence

```{r}
p_OUD_U <- (0.134 * 56.4 + 0.092 * 21.2) / (56.4 + 21.2)
p_OUD_S <- 0.045
# p_S <- 1111 / (1111 + 1478 + 538)
# p_U <- (1478 + 538) / (1111 + 1478 + 538)
```

### Excess mortality

#### All-cause mortality for OUD and unhoused populations

```{r}
mort_func <- approxfun(mortality_df$age, mortality_df$mortality, rule = 2)  # Linear interpolation

median_mortality <- mort_func(38.3)      # Median age for US population
PEH_median_mortality <- mort_func(46.5)  # Median age for PEH

U_all_mort <- 9.8 * PEH_median_mortality
S_all_mort <- U_all_mort / 2.7

YT_all_cause <- 0.0092
NT_all_cause <- 0.0489
discharged_all_cause <- 0.0169
# YT_overdose <- 0.0024
# NT_overdose <- 0.0243
discharged_overdose <- 0.0068
```

#### Excess mortality due to OUD

```{r}
YT_excess <- YT_all_cause - median_mortality
NT_excess <- NT_all_cause - median_mortality
AT_excess <- discharged_all_cause - discharged_overdose - median_mortality
```

#### Excess mortality due to homelessness

```{r}
U_excess <- U_all_mort - p_OUD_U * (p_OUD_YT * YT_excess + p_OUD_NT * NT_excess ) - PEH_median_mortality
S_excess <- S_all_mort - p_OUD_S * (p_OUD_YT * YT_excess + p_OUD_NT * NT_excess) - PEH_median_mortality
```

### Function of mortality

```{r}
getMortality <- function(age, housing, OUD) {
  background <- mort_func(age)
  annual_rate <- background + excess_matrix[cbind(housing, OUD)]
  return(rate_to_prob(annual_rate))
}
```

#### Precomputed excess mortality matrix

```{r}
excess_matrix <- matrix(c(
  0, YT_excess, NT_excess, AT_excess,                                   # H
  0, S_excess + YT_excess, S_excess + NT_excess, S_excess + AT_excess,  # S
  0, U_excess + YT_excess, U_excess + NT_excess, U_excess + AT_excess   # U
), nrow = 3, byrow = TRUE, 
dimnames = list(c("H", "S", "U"), c("NU", "YT", "NT", "AT")))
```

## State transitions

### Sampling function

```{r}
# sample_state <- function(prob_vec, states) {
#   sample(states, size = 1, prob = prob_vec)
# }
```

### Lookup tables

```{r}
housing_lookup <- setNames(1:length(v.housing), v.housing)
OUD_lookup <- setNames(1:length(v.OUD), v.OUD)
```

### Housing transitions

#### Aunnal probabilities

```{r}
# H = Permanently Housed, U = Unsheltered, S = Sheltered
H_total <- 2660 + 5121 + 2498
U_total <- 40958
S_total <- 27693 + 10730 - 4359

# Leave H
p_H_to_U <- 1148 / H_total       # Base case
#p_H_to_U <- 1148 / H_total * 2  # Sensitivity analysis
p_H_to_S <- 0

# Leave S
p_S_to_H <- (3405 + 1053 + 3876) / S_total
p_S_to_U <- (6803 + 7604 + 987 + 635) / S_total

# Leave U
p_U_to_H <-  1053 / U_total
p_U_to_S <- (6714 + 1323) / U_total
```

#### Transition probabilities for housing states

```{r}
HousingProbs <- function(current_housing, current_OUD, scenario = "base") {
  if (current_housing == "H") {
    # For permanently housed (H):
    r_H_to_S <- prob_to_rate(p_H_to_S, t = 1)
    r_H_to_U <- prob_to_rate(p_H_to_U, t = 1)
    # OUD adjustment for H→S and H→U:
    adj <- switch(current_OUD,
                  "NU" = c(S = 1, U = 1),
                  "YT" = c(S = 1, U = 1),
                  "NT" = c(S = 1, U = 1),
                  "AT" = c(S = 1, U = 1),
                  c(S = 1, U = 1))
    r_H_to_S_adj <- r_H_to_S * as.numeric(adj["S"])
    r_H_to_U_adj <- r_H_to_U * as.numeric(adj["U"])
    r_total <- r_H_to_S_adj + r_H_to_U_adj
    p_leave <- rate_to_prob(r_total, t = 1/12)
    p_H_to_S_adj <- if (r_total > 0) p_leave * (r_H_to_S_adj / r_total) else 0
    p_H_to_U_adj <- if (r_total > 0) p_leave * (r_H_to_U_adj / r_total) else 0
    p_H_to_H_adj <- 1 - p_leave
    probs <- c(H = p_H_to_H_adj, S = p_H_to_S_adj, U = p_H_to_U_adj)
    
  } else if (current_housing == "U") {
    # For Unsheltered (U):
    r_U_to_H <- prob_to_rate(p_U_to_H, t = 1)
    r_U_to_S <- prob_to_rate(p_U_to_S, t = 1)
    # OUD adjustments for U→H and U→S:
    adj <- switch(current_OUD,
                  "NU" = c(H = 1, S = 1),
                  "YT" = c(H = 1, S = 1),
                  "NT" = c(H = 1, S = 1),
                  "AT" = c(H = 1, S = 1),
                  c(H = 1, S = 1))
    multiplier_U_to_H <- ifelse(scenario == "alternative", 1.5, 1)
    multiplier_U_to_S <- 1
    r_U_to_H_adj <- r_U_to_H * multiplier_U_to_H * as.numeric(adj["H"])
    r_U_to_S_adj <- r_U_to_S * multiplier_U_to_S * as.numeric(adj["S"])
    r_total <- r_U_to_H_adj + r_U_to_S_adj
    p_leave <- rate_to_prob(r_total, t = 1/12)
    p_U_to_H_adj <- if (r_total > 0) p_leave * (r_U_to_H_adj / r_total) else 0
    p_U_to_S_adj <- if (r_total > 0) p_leave * (r_U_to_S_adj / r_total) else 0
    p_U_to_U_adj <- 1 - p_leave
    probs <- c(H = p_U_to_H_adj, S = p_U_to_S_adj, U = p_U_to_U_adj)
    
  } else if (current_housing == "S") {
    # For Sheltered (S)
    r_S_to_H <- prob_to_rate(p_S_to_H, t = 1)
    r_S_to_U <- prob_to_rate(p_S_to_U, t = 1)
    # OUD adjustments for S→H and S→U:
    adj <- switch(current_OUD,
                  "NU" = c(H = 1, U = 1),
                  "YT" = c(H = 1, U = 1),
                  "NT" = c(H = 1, U = 1),
                  "AT" = c(H = 1, U = 1),
                  c(H = 1, U = 1))
    multiplier_S_to_H <- ifelse(scenario == "alternative", 1.5, 1)
    multiplier_S_to_U <- 1
    r_S_to_H_adj <- r_S_to_H * multiplier_S_to_H * adj["H"]
    r_S_to_U_adj <- r_S_to_U * multiplier_S_to_U * adj["U"]
    r_total <- r_S_to_H_adj + r_S_to_U_adj
    p_leave <- rate_to_prob(r_total, t = 1/12)
    p_S_to_H_adj <- as.numeric(if (r_total > 0) p_leave * (r_S_to_H_adj / r_total) else 0)
    p_S_to_U_adj <- as.numeric(if (r_total > 0) p_leave * (r_S_to_U_adj / r_total) else 0)
    p_S_to_S_adj <- as.numeric(1 - p_leave)
    probs <- c(H = p_S_to_H_adj, S = p_S_to_S_adj, U = p_S_to_U_adj)
    
  } else {
    stop("Invalid housing state")
  }
  
  # Check that for the current state the sum of probabilities equals 1
  if (abs(sum(probs) - 1) > 1e-6) {
    warning("For housing state ", current_housing, ", the sum of transition probabilities is ", sum(probs))
  }
  return(probs)
}
```

#### Function for precomputing Housing Transitions

```{r}
precompute_housing_transitions <- function(scenario = "base") {
  housing_states <- c("H", "S", "U")
  OUD_states <- c("NU", "YT", "NT", "AT")
  
  # Create an empty 3D array:
  # Dimensions: current housing (rows) x current OUD (columns) x next housing (3rd dimension)
  trans_array <- array(NA, dim = c(length(housing_states), length(OUD_states), length(housing_states)),
                       dimnames = list(current_housing = housing_states,
                                       current_OUD = OUD_states,
                                       next_housing = housing_states))
  
  # Loop over all combinations of current housing and OUD states
  for (h in housing_states) {
    for (o in OUD_states) {
      # Compute the probability vector using HousingProbs function
      probs <- HousingProbs(current_housing = h, current_OUD = o, scenario = scenario)
      trans_array[h, o, ] <- probs
    }
  }
  return(trans_array)
}
```

### OUD Transitions

#### Annual rates

```{r}
r_NU_to_YT <- 0
r_NU_to_NT <- 0
r_NU_to_AT <- 0

r_YT_to_NU <- 0

p_methadone <- 356843 / (356843 + 54488)
p_buprenorphine <- 54488 / (356843 + 54488)

r_YT_to_NT <- p_methadone * 1.051 + p_buprenorphine * 1.608
r_YT_to_AT <- 0.316

r_NT_to_NU <- 0
r_NT_to_YT <- 0.426
r_NT_to_AT <- 0.0791

r_AT_to_NU <- 0
r_AT_to_YT <- 0
r_AT_to_NT <- 0.344
```

#### Transition probabilities for OUD states

```{r}
OUDProbs <- function(current_OUD, current_housing, scenario = "base") {
  if (current_OUD == "NU") {
    # For NU:
    # Transition: NU→NT.
    adj <- switch(current_housing,
                  "H" = c(NT = 1),
                  "S" = c(NT = 1),
                  "U" = c(NT = 1),
                  c(NT = 1))
    r_NU_to_NT_adj <- r_NU_to_NT * as.numeric(adj["NT"])
    r_total <- r_NU_to_NT_adj
    p_leave <- rate_to_prob(r_total, t = 1)
    p_NU_to_NT <- as.numeric(if(r_total > 0) p_leave * (r_NU_to_NT_adj / r_total) else 0)
    p_NU_to_NU <- as.numeric(1 - p_leave)
    probs <- c(NU = p_NU_to_NU, YT = 0, NT = p_NU_to_NT, AT = 0)
    
  } else if (current_OUD == "YT") {
    # For YT:
    # Transitions: YT→NT and YT→AT.
    adj <- switch(current_housing,
                  "H" = c(NT = 1, AT = 1),
                  "S" = c(NT = 1.36, AT = 0.029 / 0.316),
                  "U" = c(NT = 1.36, AT = 0.029 / 0.316),
                  c(NT = 1, AT = 1))
    multiplier_YT_to_AT <- ifelse(scenario == "alternative", 1, 1)
    adj["AT"] <- adj["AT"] * multiplier_YT_to_AT
    r_YT_to_NT_adj <- r_YT_to_NT * as.numeric(adj["NT"])
    r_YT_to_AT_adj <- r_YT_to_AT * as.numeric(adj["AT"])
    r_YT_to_NU_adj <- r_YT_to_NU
    r_total <- r_YT_to_NT_adj + r_YT_to_AT_adj + r_YT_to_NU_adj
    p_leave <- rate_to_prob(r_total, t = 1)
    p_YT_to_NT <- if(r_total > 0) p_leave * (r_YT_to_NT_adj / r_total) else 0
    p_YT_to_AT <- if(r_total > 0) p_leave * (r_YT_to_AT_adj / r_total) else 0
    p_YT_to_YT <- 1 - p_leave
    probs <- c(NU = 0, YT = p_YT_to_YT, NT = p_YT_to_NT, AT = p_YT_to_AT)
    
  } else if (current_OUD == "NT") {
    # For NT:
    # Transitions: NT→YT and NT→AT.
    adj <- switch(current_housing,
                  "H" = c(YT = 1, AT = 1),
                  "S" = c(YT = 0.213 / 0.426, AT = 0.0074 / 0.0791),
                  "U" = c(YT = 0.213 / 0.426,  AT = 0.0074 / 0.0791),
                  c(YT = 1, AT = 1))
    multiplier_NT_to_YT <- ifelse(scenario == "alternative", 1, 1)
    adj["YT"] <- adj["YT"] * multiplier_NT_to_YT
    r_NT_to_YT_adj <- r_NT_to_YT * as.numeric(adj["YT"])
    r_NT_to_AT_adj <- r_NT_to_AT * as.numeric(adj["AT"])
    r_NT_to_NU_adj <- r_NT_to_NU
    r_total <- r_NT_to_YT_adj + r_NT_to_AT_adj + r_NT_to_NU_adj
    p_leave <- rate_to_prob(r_total, t = 1)
    p_NT_to_YT <- if(r_total > 0) p_leave * (r_NT_to_YT_adj / r_total) else 0
    p_NT_to_AT <- if(r_total > 0) p_leave * (r_NT_to_AT_adj / r_total) else 0
    p_NT_to_NU <- 0
    p_NT_to_NT <- 1 - p_leave
    probs <- c(NU = 0, YT = p_NT_to_YT, NT = p_NT_to_NT, AT = p_NT_to_AT)
    
  } else if (current_OUD == "AT") {
    # For AT:
    # Transition: AT→NT.
    adj <- switch(current_housing,
                  "H" = c(NT = 1),
                  "S" = c(NT = 0.031 / 0.023),
                  "U" = c(NT = 0.031 / 0.023),
                  c(NT = 1))
    r_AT_to_NT_adj <- r_AT_to_NT * as.numeric(adj["NT"])
    r_total <- r_AT_to_NT_adj
    p_leave <- rate_to_prob(r_total, t = 1)
    p_AT_to_NT <- if(r_total > 0) p_leave * (r_AT_to_NT_adj / r_total) else 0
    p_AT_to_AT <- 1 - p_leave
    probs <- c(NU = 0, YT = 0, NT = p_AT_to_NT, AT = p_AT_to_AT)
    
  } else {
    stop("Invalid OUD state")
  }
  
  if (abs(sum(probs) - 1) > 1e-9) {
    stop("For OUD state ", current_OUD, 
         ", the sum of transition probabilities is ", sum(probs))
  }
  
  return(probs)
}
```

#### Function for precomputing OUD Transitions

```{r}
precompute_OUD_transitions <- function(scenario = "base") {
  OUD_states <- c("NU", "YT", "NT", "AT")
  housing_states <- c("H", "S", "U")
  
  # Create an empty 3D array:
  # Dimensions: current OUD (rows) x current housing (columns) x next OUD (3rd dimension)
  trans_array <- array(NA, dim = c(length(OUD_states), length(housing_states), length(OUD_states)),
                       dimnames = list(current_OUD = OUD_states,
                                       current_housing = housing_states,
                                       next_OUD = OUD_states))
  
  # Loop over all combinations of current OUD and housing states
  for (o in OUD_states) {
    for (h in housing_states) {
      # Compute the probability vector using OUDProbs function
      probs <- OUDProbs(current_OUD = o, current_housing = h, scenario = scenario)
      trans_array[o, h, ] <- probs
    }
  }
  return(trans_array)
}
```

### Precomputed transitions matrices

```{r}
housing_trans_base <- precompute_housing_transitions(scenario = "base")
housing_trans_alt  <- precompute_housing_transitions(scenario = "alternative")

OUD_trans_base <- precompute_OUD_transitions(scenario = "base")
OUD_trans_alt  <- precompute_OUD_transitions(scenario = "alternative")
```

## Costs

### Baseline costs

```{r}
c_baseline <- 2764
```

### Excess healthcare cost due to OUD

```{r}
c_health_YT <- 7052
c_health_NT <- 8827
```

### Excess healthcare cost due to homelessness

```{r}
c_health_PEH <- 13823

p_U_outdoor <- 56.4 / (56.4 + 21.2)
p_U_vehicle <- 21.2 / (56.4 + 21.2)
p_U_sick <- (p_U_outdoor * 0.6) + (p_U_vehicle * 0.43)  # Complex behavioral health needs (CASPEH unsheltered brief)
p_S_sick <- 0.25
U_S_sick_ratio <- p_U_sick / p_S_sick
S_U_sick_ratio <-  p_S_sick / p_U_sick
c_health_U <- c_health_PEH / (S_U_sick_ratio * p_S + p_U)
c_health_S <- S_U_sick_ratio * c_health_U
```

### Treatment costs

```{r}
c_treatment_YT <- p_methadone * 8584 + p_buprenorphine * 7836
```

### Overdose costs

```{r}
S_U_OD_ratio <- 281 / 78

r_OD_H_YT <- 0.0567
r_OD_H_NT <- 0.1331

c_OD_H_YT <- 3084 * r_OD_H_YT
c_OD_H_NT <- 3084 * r_OD_H_NT

r_OD_PEH_YT <- 0.0583
  
r_OD_U_YT <- r_OD_PEH_YT / (S_U_OD_ratio * p_S + p_U)
r_OD_S_YT <- S_U_OD_ratio * r_OD_U_YT

c_OD_U_YT <- 3029 * r_OD_U_YT
c_OD_S_YT <- 3029 * r_OD_S_YT

r_OD_PEH_NT <- 0.1368
  
r_OD_U_NT <- r_OD_PEH_NT / (S_U_OD_ratio * p_S + p_U)
r_OD_S_NT <- S_U_OD_ratio * r_OD_U_NT

c_OD_U_NT <- 3029 * r_OD_U_NT
c_OD_S_NT <- 3029 * r_OD_S_NT
```

### Housing costs

```{r}
c_housing_H <- 24624           # Base case
#c_housing_H <- 24624 + 23976  # Sensitivity analysis
p_emergency_shelter <- 0.15 / p_S
p_transitional_shelter <- 1 - p_emergency_shelter
c_emergency_shelter <- 18283
c_transitional_shelter <- 35408

c_shelter_S <- p_emergency_shelter * c_emergency_shelter + p_transitional_shelter * c_transitional_shelter
```

### Cost estimates

```{r}
c_H_YT <- c_baseline + c_health_YT + c_treatment_YT + c_OD_H_YT + c_housing_H
c_H_NT <- c_baseline + c_health_NT + c_OD_H_NT + c_housing_H
c_H_AT <- c_baseline + c_health_YT + c_housing_H
c_S_YT <- c_baseline + c_health_YT + c_treatment_YT + c_health_S + c_OD_S_YT + c_shelter_S
c_S_NT <- c_baseline + c_health_NT + c_health_S + c_OD_S_NT + c_shelter_S
c_S_AT <- c_baseline + c_health_YT + c_health_S + c_shelter_S
c_U_YT <- c_baseline + c_health_U + c_health_YT + c_treatment_YT + c_OD_U_YT
c_U_NT <- c_baseline + c_health_NT + c_health_U + c_OD_U_NT
c_U_AT <- c_baseline + c_health_YT + c_health_U
```

### Precomputed cost matrix

```{r}
cost_matrix <- matrix(c(
  0, c_H_YT, c_H_NT, c_H_AT,   # H
  0, c_S_YT, c_S_NT, c_S_AT,   # S
  0, c_U_YT, c_U_NT, c_U_AT    # U
), nrow = 3, byrow = TRUE, 
dimnames = list(c("H", "S", "U"), c("NU", "YT", "NT", "AT")))
```

## QALYs

### QALYs estimates

```{r}
q_PEH <- 0.43   # Base case
#q_PEH <- 0.60  # Sensitivity analysis
q_U <- q_PEH / (U_S_sick_ratio * p_S + p_U)
q_S <- U_S_sick_ratio * q_U
q_YT <- 0.721
q_NT <- 0.668
q_AT <- 0.80
```

### Precomupted QALYs matrix

```{r}
qalys_matrix <- matrix(c(
  1, q_YT, q_NT, q_AT,   # H
  1, q_YT * q_S, q_NT * q_S, q_AT * q_S,   # S
  1, q_YT * q_U, q_NT * q_U, q_AT * q_U    # U
), nrow = 3, byrow = TRUE, 
dimnames = list(c("H", "S", "U"), c("NU", "YT", "NT", "AT")))
```

# Microsimulation

## Function for vectorizing

```{r}
library(matrixStats)

vectorized_sample <- function(probs_mat) {
  # Each row of probs_mat is a probability vector that sums to 1
  u <- runif(nrow(probs_mat))
  # Compute cumulative sums row-wise
  cumprobs <- rowCumsums(probs_mat)
  # Determine the first column where cumulative probability >= u for each trow
  state_index <- max.col(cumprobs >= u, ties.method = "first")
  if (!is.null(colnames(probs_mat))) {
    return(colnames(probs_mat)[state_index])
  } else {
    return(state_index)
  }
}
```

## Function of performing microsimulation

```{r}
MicroSim_complex <- function(n.i, n.t, init_age, init_sex, init_housing, init_OUD, d.c, d.e, scenario = "base", seed = 1) {
  set.seed(seed)
  
  if (scenario == "base"){
    housing_trans <- housing_trans_base
    OUD_trans <- OUD_trans_base
  } else {
    housing_trans <- housing_trans_alt
    OUD_trans <- OUD_trans_alt
  }
  
  id <- 1:n.i
  
  m.age     <- matrix(NA, nrow = n.i, ncol = n.t + 1)
  m.sex     <- matrix(NA, nrow = n.i, ncol = n.t + 1)
  m.housing <- matrix(NA, nrow = n.i, ncol = n.t + 1)
  m.OUD     <- matrix(NA, nrow = n.i, ncol = n.t + 1)
  m.dead    <- matrix(NA, nrow = n.i, ncol = n.t + 1)
  m.cost    <- matrix(0, nrow = n.i, ncol = n.t + 1)
  m.QALY    <- matrix(0, nrow = n.i, ncol = n.t + 1)
  
  # Initialize cycle 0
  m.age[, 1]     <- init_age
  m.sex[, 1]     <- init_sex
  m.housing[, 1] <- init_housing
  m.OUD[, 1]     <- init_OUD
  m.dead[, 1]    <- init_dead
  
  m.cost[, 1] <- cost_matrix[cbind(m.housing[, 1], m.OUD[, 1])] * (1/12)
  m.QALY[, 1] <- qalys_matrix[cbind(m.housing[, 1], m.OUD[, 1])] * (1/12)
  
  # Discount vectors (for cycles 0 through n.t).
  v.dwc <- 1 / (1 + d.c)^(0:n.t)
  v.dwe <- 1 / (1 + d.e)^(0:n.t)
  
  ##################################################
  # Simulation Loop: Cycle-by-Cycle Update
  ##################################################
  
  for (t in 1:n.t) {
    # Identify indices for alive and dead individuals at cycle t
    alive <- which(m.dead[, t] == 0)
    dead  <- which(m.dead[, t] == 1)
    
    # For individuals already dead: Copy forward their state
    if (length(dead) > 0) {
      m.age[dead, t + 1]     <- m.age[dead, t]
      m.sex[dead, t + 1]     <- m.sex[dead, t]
      m.housing[dead, t + 1] <- m.housing[dead, t]
      m.OUD[dead, t + 1]     <- m.OUD[dead, t]
      m.dead[dead, t + 1]    <- 1
      m.cost[dead, t + 1]    <- 0
      m.QALY[dead, t + 1]    <- 0
    }
    
    # For alive individuals:
    if (length(alive) > 0) {
      # Update age and sex
      new_age <- m.age[alive, t] + 1/12
      m.age[alive, t + 1] <- new_age
      m.sex[alive, t + 1] <- m.sex[alive, t]
      
      # Compute mortality
      p_death <- getMortality(new_age, m.housing[alive, t], m.OUD[alive, t])
      
      # Compute death events
      death_event <- rbinom(length(alive), 1, p_death)
      
      # For those who died:
      died_idx <- alive[which(death_event == 1)]
      if (length(died_idx) > 0) {
        m.housing[died_idx, t + 1] <- m.housing[died_idx, t]
        m.OUD[died_idx, t + 1]     <- m.OUD[died_idx, t]
        m.dead[died_idx, t + 1]    <- 1
        m.cost[died_idx, t + 1]    <- 0
        m.QALY[died_idx, t + 1]    <- 0
      }
      
      # For survivors
      # Update housing and OUD transitions
      surv_idx <- alive[which(death_event == 0)]

      if (length(surv_idx) > 0) {
      # Extract current states for survivors.
      curr_housing <- m.housing[surv_idx, t]
      curr_OUD     <- m.OUD[surv_idx, t]
      
      
      # Convert to numeric indices.
        curr_housing_idx <- housing_lookup[curr_housing]
        curr_OUD_idx     <- OUD_lookup[curr_OUD]
        n_surv <- length(surv_idx)
        
        # Vectorized extraction for Housing Transitions
        # Dimensions of housing_trans: [current housing x current OUD x next housing]
        dim1 <- dim(housing_trans)[1]  # 3
        dim2 <- dim(housing_trans)[2]  # 4
        n_next <- length(v.housing)    # 3
        # Compute linear indices:
        indices_housing <- sapply(1:n_next, function(j) {
          curr_housing_idx + (curr_OUD_idx - 1) * dim1 + (j - 1) * dim1 * dim2
        })
        probs_housing <- housing_trans[as.vector(indices_housing)]
        dim(probs_housing) <- c(n_surv, n_next)
        colnames(probs_housing) <- v.housing
        m.housing[surv_idx, t + 1] <- vectorized_sample(probs_housing)
        
        # Vectorized extraction for OUD Transitions
        # Dimensions of OUD_trans: [current OUD x current housing x next OUD]
        dim1_OUD <- 4   # dim(OUD_trans)[1]  # 4
        dim2_OUD <- 3   # dim(OUD_trans)[2]  # 3
        n_next_OUD <- 4 # length(v.OUD)      # 4
        indices_OUD <- sapply(1:n_next_OUD, function(j) {
          curr_OUD_idx + (curr_housing_idx - 1) * dim1_OUD + (j - 1) * dim1_OUD * dim2_OUD
        })
        probs_OUD <- OUD_trans[as.vector(indices_OUD)]
        dim(probs_OUD) <- c(n_surv, n_next_OUD)
        colnames(probs_OUD) <- v.OUD
        m.OUD[surv_idx, t + 1] <- vectorized_sample(probs_OUD)
        
        m.dead[surv_idx, t + 1] <- 0
 # Update cost and QALY using lookup tables
        new_housing <- m.housing[surv_idx, t + 1]
        new_OUD     <- m.OUD[surv_idx, t + 1]
        m.cost[surv_idx, t + 1] <- cost_matrix[cbind(new_housing, new_OUD)] * (1/12)
        m.QALY[surv_idx, t + 1] <- qalys_matrix[cbind(new_housing, new_OUD)] * (1/12)
      }
    }
    #cat('\r', paste(round(t / n.t * 100), "% done", sep = " "))
  }
  #cat("\nSimulation complete.\n")
  
  
  # Compute total (discounted) cost and QALYs for each individual.
  tc <- m.cost %*% v.dwc
  tq <- m.QALY %*% v.dwe
  
  #tc_hat <- mean(tc)
  #tq_hat <- mean(tq)
  #mcse_cost <- sd(tc) / sqrt(n.i)
  #mcse_qaly <- sd(tq) / sqrt(n.i)
  
  results <- list(ID = id, age = m.age, sex = m.sex, housing = m.housing, OUD = m.OUD, dead = m.dead, 
                  cost = tc, QALY = tq
                  #cost = m.cost, QALY = m.QALY,
                  #total_cost = tc, total_QALY = tq,
                  #avg_cost = tc_hat, avg_QALY = tq_hat,
                  #mcse_cost = mcse_cost, mcse_qaly = mcse_qaly
                  )
  return(results)
}
```

## Function of saving results for each iteration

```{r}
library(data.table)

sim_to_dt <- function(sim, scenario_label, iteration, n_cycles, n.i) {
  dt <- data.table(
    iteration = rep(iteration, each = n.i * n_cycles),
    scenario  = scenario_label,
    ID        = rep(sim$ID, each = n_cycles),
    Cycle     = rep(0:(n_cycles - 1), times = n.i),
    Age       = as.vector(t(sim$age)),
    Sex       = as.vector(t(sim$sex)),
    Housing   = as.vector(t(sim$housing)),
    OUD       = as.vector(t(sim$OUD)),
    Dead      = as.vector(t(sim$dead)),
    Cost      = as.vector(t(sim$cost)),
    QALY      = as.vector(t(sim$QALY))
  )
  return(dt)
}

# Cycle summary
create_cycle_summary <- function(sim, scenario_label, iteration, n_cycles, n.i) {
  # Convert simulation result to long format and then to data.table
  dt <- sim_to_dt(sim, scenario_label, iteration, n_cycles, n.i)
  
  # Compute the summary by grouping on iteration and Cycle.
  cycle_summary <- dt[, .(
    count_H    = sum(Dead == 0 & Housing == "H"),
    count_U    = sum(Dead == 0 & Housing == "U"),
    count_S    = sum(Dead == 0 & Housing == "S"),
    count_NT   = sum(Dead == 0 & OUD == "NT"),
    count_YT   = sum(Dead == 0 & OUD == "YT"),
    count_AT   = sum(Dead == 0 & OUD == "AT"),
    count_H_NT = sum(Dead == 0 & Housing == "H" & OUD == "NT"),
    count_H_YT = sum(Dead == 0 & Housing == "H" & OUD == "YT"),
    count_H_AT = sum(Dead == 0 & Housing == "H" & OUD == "AT"),
    count_U_NT = sum(Dead == 0 & Housing == "U" & OUD == "NT"),
    count_U_YT = sum(Dead == 0 & Housing == "U" & OUD == "YT"),
    count_U_AT = sum(Dead == 0 & Housing == "U" & OUD == "AT"),
    count_S_NT = sum(Dead == 0 & Housing == "S" & OUD == "NT"),
    count_S_YT = sum(Dead == 0 & Housing == "S" & OUD == "YT"),
    count_S_AT = sum(Dead == 0 & Housing == "S" & OUD == "AT"),
    count_Dead = sum(Dead == 1),
    total_cost = sum(Cost),
    total_QALY = sum(QALY)
  ), by = .(iteration, Cycle, scenario)]
  
  return(cycle_summary)
}
```

## Performing microsimulation

```{r}
#library(parallel)

n_iter <- 1000
#num_cores <- detectCores() - 1

overall_start <- Sys.time()

results_list <- mclapply(1:n_iter, function(iter) {
  # Run base scenario
  sim_base <- MicroSim_complex(n.i = n.i, n.t = n.t, init_age = init_age, init_sex = init_sex,
                               init_housing = init_housing, init_OUD = init_OUD,
                               d.c = d.c, d.e = d.e, scenario = "base", seed = 100 + iter)
  
  # Run alternative scenario
  sim_alt <- MicroSim_complex(n.i = n.i, n.t = n.t, init_age = init_age, init_sex = init_sex,
                              init_housing = init_housing, init_OUD = init_OUD,
                              d.c = d.c, d.e = d.e, scenario = "alternative", seed = 100 + iter)
  
  # Create cycle summaries for both scenarios.
  sim_base_summary <- create_cycle_summary(sim_base, scenario_label = "base",
                                           iteration = iter, n_cycles = n.t + 1, n.i = n.i)
  sim_alt_summary <- create_cycle_summary(sim_alt, scenario_label = "alternative",
                                          iteration = iter, n_cycles = n.t + 1, n.i = n.i)
  
  #cat("Iteration", iter, "completed.\n")
  
  list(sim_base_summary = sim_base_summary, sim_alt_summary = sim_alt_summary)
}
  #, mc.cores = num_cores
  )

overall_end <- Sys.time()
overall_time <- overall_end - overall_start
cat("Overall computation time for", n_iter, "iterations:",
    as.numeric(overall_time, units = "mins"), "minutes.\n")

# Extract the lists of cycle summaries for each scenario.
sim_results_base <- lapply(results_list, function(x) x$sim_base_summary)
sim_results_alt  <- lapply(results_list, function(x) x$sim_alt_summary)
```

## Export results

```{r}
# Base case
#saveRDS(sim_results_base, "~/Documents/PEH-OUD/sim_results_base.rds")
#saveRDS(sim_results_alt, "~/Documents/PEH-OUD/sim_results_alt.rds")

# Sensitivity analysis: Housing cost 2x
#saveRDS(sim_results_base, "~/Documents/PEH-OUD/sim_results_base_housed_cost_x2.rds")
#saveRDS(sim_results_alt, "~/Documents/PEH-OUD/sim_results_alt_housed_cost_x2.rds")

# Sensitivity analysis: Rate to unhoused 2x
#saveRDS(sim_results_base, "~/Documents/PEH-OUD/sim_results_base_rate_unhoused_x2.rds")
#saveRDS(sim_results_alt, "~/Documents/PEH-OUD/sim_results_alt_rate_unhoused_x2.rds")

# Sensitivity analysis: QALY multiplier 40% increase
#saveRDS(sim_results_base, "~/Documents/PEH-OUD/sim_results_base_change_qaly.rds")
#saveRDS(sim_results_alt, "~/Documents/PEH-OUD/sim_results_alt_change_qaly.rds")
```

# CEA Analysis

## Read in the simulation result files for each case

```{r}
path <- "~/Documents/PEH-OUD"

sim_results_base               <- readRDS(file.path(path, "sim_results_base.rds"))
sim_results_alt                <- readRDS(file.path(path, "sim_results_alt.rds"))
sim_results_base_housing_cost  <- readRDS(file.path(path, "sim_results_base_housing_cost_x2.rds"))
sim_results_alt_housing_cost   <- readRDS(file.path(path, "sim_results_alt_housing_cost_x2.rds"))
sim_results_base_rate_unhoused <- readRDS(file.path(path, "sim_results_base_rate_unhoused_x2.rds"))
sim_results_alt_rate_unhoused  <- readRDS(file.path(path, "sim_results_alt_rate_unhoused_x2.rds"))
sim_results_base_change_qaly   <- readRDS(file.path(path, "sim_results_base_change_qaly.rds"))
sim_results_alt_change_qaly    <- readRDS(file.path(path, "sim_results_alt_change_qaly.rds"))
```

## Function of computing the per-iteration summary

```{r}
n.i <- 1000  # number of simulated individuals

per_iteration_summary <- function(sim_base_list, sim_alt_list, n.i) {
  iteration_summary <- lapply(seq_along(sim_base_list), function(i) {
    base_df <- sim_base_list[[i]]
    alt_df  <- sim_alt_list[[i]]
    
    # Extract overall (discounted) cost and QALY and compute per-individual values
    total_cost_base <- base_df$total_cost[1] / n.i
    total_QALY_base <- base_df$total_QALY[1] / n.i
    total_cost_alt  <- alt_df$total_cost[1] / n.i
    total_QALY_alt  <- alt_df$total_QALY[1] / n.i
    
    delta_cost <- total_cost_alt - total_cost_base
    delta_QALY <- total_QALY_alt - total_QALY_base
    ICER <- ifelse(delta_QALY != 0, delta_cost / delta_QALY, NA)
    
    data.frame(
      iteration = i,
      total_cost_base = total_cost_base,
      total_QALY_base = total_QALY_base,
      total_cost_alt  = total_cost_alt,
      total_QALY_alt  = total_QALY_alt,
      delta_cost      = delta_cost,
      delta_QALY      = delta_QALY,
      ICER            = ICER
    )
  })
  do.call(rbind, iteration_summary)
}
```

## Compute per-iteration summaries for each case

```{r}
library(dplyr)

iter_sum_base_case <- per_iteration_summary(sim_results_base, sim_results_alt, n.i) %>% 
  mutate(case = "Base case")
iter_sum_housing_cost <- per_iteration_summary(sim_results_base_housing_cost,
                                               sim_results_alt_housing_cost, n.i) %>% 
  mutate(case = "Housing cost x2 (no rent)")
iter_sum_rate_unhoused <- per_iteration_summary(sim_results_base_rate_unhoused,
                                                sim_results_alt_rate_unhoused, n.i) %>% 
  mutate(case = "Housed to unhoused rate x2")

iter_sum_change_qaly <- per_iteration_summary(sim_results_base_change_qaly,
                                              sim_results_alt_change_qaly, n.i) %>% 
  mutate(case = "40% increase in PEH QoL")

# Combine all per-iteration summaries
iteration_summary_df <- bind_rows(iter_sum_base_case,
                                  iter_sum_housing_cost,
                                  iter_sum_rate_unhoused,
                                  iter_sum_change_qaly)

overall_summary <- iteration_summary_df %>%
  group_by(case) %>%
  summarise(
    mean_total_cost_base = mean(total_cost_base),
    mean_total_QALY_base = mean(total_QALY_base),
    mean_total_cost_alt  = mean(total_cost_alt),
    mean_total_QALY_alt  = mean(total_QALY_alt),
    mean_delta_cost      = mean(delta_cost),
    mean_delta_QALY      = mean(delta_QALY),
    mean_ICER            = mean(ICER, na.rm = TRUE),
    mcse_total_cost_base = sd(total_cost_base) / sqrt(n()),
    mcse_total_QALY_base = sd(total_QALY_base) / sqrt(n()),
    mcse_total_cost_alt  = sd(total_cost_alt) / sqrt(n()),
    mcse_total_QALY_alt  = sd(total_QALY_alt) / sqrt(n()),
    mcse_delta_cost      = sd(delta_cost) / sqrt(n()),
    mcse_delta_QALY      = sd(delta_QALY) / sqrt(n()),
    mcse_ICER            = sd(ICER, na.rm = TRUE) / sqrt(n())
  )
```

# Decision table

```{r}
library(tibble)
library(dplyr)

overall_summary <- overall_summary %>%
  mutate(case = factor(case, levels = c("Base case", 
                                        "Housing cost x2 (no rent)",
                                        "Housed to unhoused rate x2",
                                        "40% increase in PEH QoL")))

decision_table <- overall_summary %>%
  arrange(case) %>%
  group_by(case) %>%
  do(
    tibble(
      case = unique(.$case),
      Strategy = c("Status Quo", "Housing Intervention"),
      Cost = c(.$mean_total_cost_base, .$mean_total_cost_alt),
      QALY = c(.$mean_total_QALY_base, .$mean_total_QALY_alt),
      Inc_Cost = c(NA, .$mean_delta_cost),
      Inc_QALY = c(NA, .$mean_delta_QALY),
      ICER = c(NA, .$mean_ICER)
    )
  ) %>%
  ungroup() %>%
  mutate(
    Cost = round(Cost, 0),
    QALY = round(QALY, 2),
    Inc_Cost = round(Inc_Cost, 0),
    Inc_QALY = round(Inc_QALY, 2),
    ICER = round(ICER, 0)
  )

print(decision_table)

# decision_table <- decision_table %>%
#   mutate(
#     Cost = format(round(as.numeric(Cost), 0), big.mark = ","),
#     QALY = format(round(as.numeric(QALY), 2), nsmall = 2),
#     Inc_Cost = ifelse(is.na(Inc_Cost), NA, format(round(as.numeric(Inc_Cost), 0), big.mark = ",")),
#     Inc_QALY = ifelse(is.na(Inc_QALY), NA, format(round(as.numeric(Inc_QALY), 2), nsmall = 2)),
#     ICER = ifelse(is.na(ICER), NA, format(round(as.numeric(ICER), 0), big.mark = ","))
#   )
# 
# write.csv(decision_table, "~/Documents/GitHub/PEH-OUD/decision_table.csv")
```

# Figures

## CE plot

```{r}
library(ggplot2)
library(ggrepel)
library(scales)

plot_icers <- function(data, axis_text_size = 12, axis_title_size = 14) {
  ggplot(data, aes(x = Cost, y = QALY, label = Strategy)) +
    geom_point(size = 3) +
    geom_line(aes(group = case), linetype = "solid", color = "black") +
    geom_label_repel(nudge_x = -500, nudge_y = 0.025, size = 3, max.overlaps = Inf, segment.color = NA) +
    labs(x = "Cost (2024 USD)", y = "QALYs") +
    scale_x_continuous(labels = comma, limits = c(362500, 375000), breaks = seq(360000, 375000, by = 2500)) +
    scale_y_continuous(limits = c(4.1, 4.6), breaks = seq(4.1, 4.6, by = 0.1)) +
    theme_minimal(base_size = axis_text_size) +
    theme(
      axis.title = element_text(size = 10, face = "bold"),
      axis.text = element_text(size = 10),
      panel.background = element_rect(fill = "transparent", color = NA),
      plot.background = element_rect(fill = "transparent", color = NA),
      panel.grid.major = element_line(color = "gray80"),
      panel.grid.minor = element_line(color = "gray90")
    )
}

icer_plot_base_case <- plot_icers(decision_table %>% filter(case == "Base case"))
print(icer_plot_base_case)
```

## Bar plot for sensitivity analyses results

### Statistics calculation

```{r}
library(tidyr)

plot_data <- overall_summary %>%
  rename(
    mean_cost = mean_delta_cost,
    mcse_cost = mcse_delta_cost,
    mean_qaly = mean_delta_QALY,
    mcse_qaly = mcse_delta_QALY
  ) %>%
  mutate(
    lower_cost = mean_cost - 1.96 * mcse_cost,
    upper_cost = mean_cost + 1.96 * mcse_cost,
    lower_qaly = mean_qaly - 1.96 * mcse_qaly,
    upper_qaly = mean_qaly + 1.96 * mcse_qaly,
    lower_icer = ifelse(is.na(mean_ICER), NA, mean_ICER - 1.96 * mcse_ICER),
    upper_icer = ifelse(is.na(mean_ICER), NA, mean_ICER + 1.96 * mcse_ICER)
  )

long_df <- plot_data %>%
  pivot_longer(
    cols = c(mean_cost, lower_cost, upper_cost,
             mean_qaly, lower_qaly, upper_qaly,
             mean_ICER, lower_icer, upper_icer),
    names_to = c("stat", "Outcome"),
    names_sep = "_"  # splits e.g. "mean_cost" into stat = "mean" and Outcome = "cost"
  ) %>%
  pivot_wider(names_from = stat, values_from = value) %>%
  mutate(
    Outcome = recode(tolower(Outcome),
                     cost = "Incremental Cost",
                     qaly = "Incremental QALYs",
                     icer = "ICER"),
    Outcome = factor(Outcome, levels = c("Incremental Cost", "Incremental QALYs", "ICER"))
  )
```

## Bar charts with error bars

```{r}
ggplot(long_df, aes(x = case, y = mean, fill = case)) +
  geom_bar(stat = "identity", width = 0.7) +
  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.2) +
  facet_wrap(~ Outcome, scales = "free_y") +
  labs(x = "Case", y = "Mean Value (95% CIs)") +
  scale_x_discrete() +
  scale_y_continuous(
    limits = c(0, NA),
    breaks = pretty_breaks(n = 5),
    labels = comma
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 45, hjust = 1, size = 10, color = "black"),
    axis.text.y = element_text(size = 10),
    axis.title.x = element_text(size = 10, face = "bold"),
    axis.title.y = element_text(size = 10, face = "bold"),
    panel.background = element_rect(fill = "transparent", color = NA),
    plot.background = element_rect(fill = "transparent", color = NA),
    #panel.grid.major = element_line(color = "gray80"),
    #panel.grid.minor = element_line(color = "gray90")
  )
```

## Trajectories and prevalence

### Combining results

```{r}
dt_base <- rbindlist(sim_results_base)
dt_alt  <- rbindlist(sim_results_alt)
dt_all  <- rbind(dt_base, dt_alt)
```

### Housing states

#### Statistics calculation

```{r}
n_iter = 1000

# Reshape data to long format for housing state counts
dt_long_housing <- melt(dt_all,
                        id.vars = c("iteration", "scenario", "Cycle", "count_Dead"),
                        measure.vars = c("count_H", "count_S", "count_U", "count_Dead"),
                        variable.name = "HousingState",
                        value.name = "Count")
dt_long_housing[, Survivors := n.i - count_Dead]
dt_long_housing[, PropAll := Count / n.i]         # Proportion among all individuals
dt_long_housing[, PropSurv := Count / Survivors]  # Proportion among survivors
dt_long_housing[, HousingState := factor(HousingState,
                                         levels = c("count_H", "count_S", "count_U", "count_Dead"),
                                         labels = c("Housed", "Sheltered", "Unsheltered", "Dead"))]

# Compute summary statistics
dt_stats_housing <- dt_long_housing[, .(
  MeanProp = mean(PropSurv),
  SDProp   = sd(PropSurv),
  MeanCount = mean(Count) / n.i,
  SDCount   = sd(Count),
  LowerProp = quantile(PropSurv, probs = 0.025),
  UpperProp = quantile(PropSurv, probs = 0.975),
  LowerCount = quantile(Count, probs = 0.025) / n.i,
  UpperCount = quantile(Count, probs = 0.975) / n.i
), by = .(scenario, Cycle, HousingState)]

dt_stats_housing[, LowerBoundProp := MeanProp - 1.96 * SDProp / sqrt(n_iter)]
dt_stats_housing[, UpperBoundProp := MeanProp + 1.96 * SDProp / sqrt(n_iter)]
dt_stats_housing[, LowerBoundCount := MeanCount - 1.96 * SDCount / sqrt(n_iter)]
dt_stats_housing[, UpperBoundCount := MeanCount + 1.96 * SDCount / sqrt(n_iter)]

dt_stats_housing[, Year := Cycle / 12]
dt_stats_housing[, scenario := factor(scenario, levels = c("base", "alternative"),
                                      labels = c("Status Quo", "Housing Intervention"))]
```

#### Trajectory

```{r}
# Create the plot
p_housing_trajectory <- ggplot(dt_stats_housing, aes(x = Year, y = MeanCount, color = HousingState, fill = HousingState)) +
  geom_ribbon(aes(ymin = LowerBoundCount, ymax = UpperBoundCount, group = HousingState),
              alpha = 0.3, color = NA) +
  geom_line(size = 1) +
  labs(title = "Trajectory of Housing States Over Time",
       x = "Year",
       y = "Proportion",
       color = "Housing States",
       fill = "Housing States") +
  facet_wrap(~ scenario, ncol = 2) +
  scale_x_continuous(limits = c(0, 50), breaks = seq(0, 50, by = 10)) +
  scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, by = 0.1)) +
  # scale_color_manual(values = c("Housed" = "darkblue", 
  #                               "Sheltered" = "orange", 
  #                               "Unsheltered" = "darkgreen")) +
  # scale_fill_manual(values = c("Housed" = "lightblue", 
  #                              "Sheltered" = "yellow", 
  #                              "Unsheltered" = "lightgreen")) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5),  # Title font size
    axis.title = element_text(size = 10, face = "bold"),    # X and Y axis labels
    axis.text = element_text(size = 10),                    # X and Y axis tick labels
    legend.title = element_text(size = 10, face = "bold"),  # Legend title size
    legend.text = element_text(size = 10),                  # Legend item labels size
    strip.text = element_text(size = 12, face = "bold"),    # Facet title size
    panel.background = element_rect(fill = "transparent", color = NA),
    plot.background  = element_rect(fill = "transparent", color = NA),
    legend.position  = "top"
    # panel.grid.major = element_line(color = "grey80"),
    # panel.grid.minor = element_line(color = "grey90")
    # panel.grid.minor = element_blank()
  )

# Display the plot
print(p_housing_trajectory)
```

#### Prevalence

```{r}
p_housing_prevalence <- ggplot(dt_stats_housing[HousingState != "Dead", ],
                               aes(x = Year, y = MeanProp, color = HousingState, fill = HousingState)) +
  geom_ribbon(aes(ymin = LowerBoundProp, ymax = UpperBoundProp, group = HousingState),
              alpha = 0.3, color = NA) +
  geom_line(size = 1) +
  labs(title = "Prevalence of Housing States Over Time",
       x = "Year",
       y = "Proportion",
       color = "Housing States",
       fill = "Housing States") +
  facet_wrap(~ scenario, ncol = 2) +
  scale_x_continuous(limits = c(0, 50), breaks = seq(0, 50, by = 10)) +
  scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, by = 0.1)) +
  # scale_color_manual(values = c("Housed" = "darkblue", 
  #                               "Sheltered" = "orange", 
  #                               "Unsheltered" = "darkgreen")) +
  # scale_fill_manual(values = c("Housed" = "lightblue", 
  #                              "Sheltered" = "yellow", 
  #                              "Unsheltered" = "lightgreen")) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
    axis.title = element_text(size = 10, face = "bold"),
    axis.text = element_text(size = 10),
    legend.title = element_text(size = 10, face = "bold"),
    legend.text = element_text(size = 10),
    strip.text = element_text(size = 12, face = "bold"),
    panel.background = element_rect(fill = "transparent", color = NA),
    plot.background  = element_rect(fill = "transparent", color = NA),
    legend.position  = "top"
    # panel.grid.major = element_line(color = "grey80"),
    # panel.grid.minor = element_line(color = "grey90")
    # panel.grid.minor = element_blank()
  )

print(p_housing_prevalence)
```

### OUD states

#### Statistics calculation

```{r}
dt_long_OUD <- melt(dt_all,
                    id.vars = c("iteration", "scenario", "Cycle", "count_Dead"),
                    measure.vars = c("count_AT", "count_YT", "count_NT", "count_Dead"),
                    variable.name = "OUDState",
                    value.name = "Count")

dt_long_OUD[, Survivors := n.i - count_Dead]
dt_long_OUD[, PropAll := Count / n.i]
dt_long_OUD[, PropSurv := Count / Survivors]
dt_long_OUD[, OUDState := factor(OUDState,
                                 levels = c("count_AT", "count_YT", "count_NT", "count_Dead"),
                                 labels = c("Abstinent", "OUD (In Treatment)", "OUD (Out of Treatment)", "Dead"))]

# Compute summary statistics
dt_stats_OUD <- dt_long_OUD[, .(
  MeanProp = mean(PropSurv),
  SDProp   = sd(PropSurv),
  MeanCount = mean(Count) / n.i,
  SDCount   = sd(Count),
  LowerProp = quantile(PropSurv, probs = 0.025),
  UpperProp = quantile(PropSurv, probs = 0.975),
  LowerCount = quantile(Count, probs = 0.025) / n.i,
  UpperCount = quantile(Count, probs = 0.975) / n.i
), by = .(scenario, Cycle, OUDState)]

dt_stats_OUD[, LowerBoundProp := MeanProp - 1.96 * SDProp / sqrt(n_iter)]
dt_stats_OUD[, UpperBoundProp := MeanProp + 1.96 * SDProp / sqrt(n_iter)]
dt_stats_OUD[, LowerBoundCount := MeanCount - 1.96 * SDCount / sqrt(n_iter)]
dt_stats_OUD[, UpperBoundCount := MeanCount + 1.96 * SDCount / sqrt(n_iter)]

dt_stats_OUD[, Year := Cycle / 12]
dt_stats_OUD[, scenario := factor(scenario, levels = c("base", "alternative"),
                                  labels = c("Status Quo", "Housing Intervention"))]
```

#### Trajectory

```{r}
p_OUD_trajectory <- ggplot(dt_stats_OUD, aes(x = Year, y = MeanCount, color = OUDState, fill = OUDState)) +
  geom_ribbon(aes(ymin = LowerBoundCount, ymax = UpperBoundCount, group = OUDState),
              alpha = 0.3, color = NA) +
  geom_line(size = 1) +
  labs(title = "Trajectory of OUD States Over Time",
       x = "Year",
       y = "Proportion",
       color = "Housing States",
       fill = "Housing States") +
  facet_wrap(~ scenario, ncol = 2) +
  scale_x_continuous(limits = c(0, 50), breaks = seq(0, 50, by = 10)) +
  scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, by = 0.1)) +
  # scale_color_manual(values = c("Housed" = "darkblue", 
  #                               "Sheltered" = "orange", 
  #                               "Unsheltered" = "darkgreen")) +
  # scale_fill_manual(values = c("Housed" = "lightblue", 
  #                              "Sheltered" = "yellow", 
  #                              "Unsheltered" = "lightgreen")) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
    axis.title = element_text(size = 10, face = "bold"),
    axis.text = element_text(size = 10),
    legend.title = element_text(size = 10, face = "bold"),
    legend.text = element_text(size = 10),
    strip.text = element_text(size = 12, face = "bold"),
    panel.background = element_rect(fill = "transparent", color = NA),
    plot.background  = element_rect(fill = "transparent", color = NA),
    legend.position  = "top"
    # panel.grid.major = element_line(color = "grey80"),
    # panel.grid.minor = element_line(color = "grey90")
    # panel.grid.minor = element_blank()
  )

print(p_OUD_trajectory)
```

#### Prevalence

```{r}
p_housing_prevalence <- ggplot(dt_stats_OUD[OUDState != "Dead", ],
                               aes(x = Year, y = MeanProp, color = OUDState, fill = OUDState)) +
  geom_ribbon(aes(ymin = LowerBoundProp, ymax = UpperBoundProp, group = OUDState),
              alpha = 0.3, color = NA) +
  geom_line(size = 1) +
  labs(title = "Prevalence of OUD States Over Time",
       x = "Year",
       y = "Proportion",
       color = "Housing States",
       fill = "Housing States") +
  facet_wrap(~ scenario, ncol = 2) +
  scale_x_continuous(limits = c(0, 50), breaks = seq(0, 50, by = 10)) +
  scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, by = 0.1)) +
  # scale_color_manual(values = c("Housed" = "darkblue", 
  #                               "Sheltered" = "orange", 
  #                               "Unsheltered" = "darkgreen")) +
  # scale_fill_manual(values = c("Housed" = "lightblue", 
  #                              "Sheltered" = "yellow", 
  #                              "Unsheltered" = "lightgreen")) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
    axis.title = element_text(size = 10, face = "bold"),
    axis.text = element_text(size = 10),
    legend.title = element_text(size = 10, face = "bold"),
    legend.text = element_text(size = 10),
    strip.text = element_text(size = 12, face = "bold"),
    panel.background = element_rect(fill = "transparent", color = NA),
    plot.background  = element_rect(fill = "transparent", color = NA),
    legend.position  = "top"
    # panel.grid.major = element_line(color = "grey80"),
    # panel.grid.minor = element_line(color = "grey90")
    # panel.grid.minor = element_blank()
  )

print(p_housing_prevalence)
```

## Raincloud plots

### Total costs

```{r}
iter_sum_base_case <- per_iteration_summary(sim_results_base, sim_results_alt, n_i) %>% 
  mutate(case = "Base case")
dt_iter <- as.data.table(iter_sum_base_case)

library(gghalves)

df_long_cost <- melt(
  dt_iter, 
  id.vars = "iteration", 
  measure.vars = c("total_cost_base", "total_cost_alt"),
  variable.name = "scenario", 
  value.name = "total_cost"
)

df_long_cost[, scenario := factor(
  scenario, 
  levels = c("total_cost_base", "total_cost_alt"),
  labels = c("Status Quo", "Housing Intervention")
)]

# Create a half‐violin (raincloud) plot for total costs
p_cost <- ggplot(df_long_cost, aes(x = scenario, y = total_cost, fill = scenario)) +
  geom_half_violin(side = "r", alpha = 0.6, trim = TRUE, color = NA) +           # Half-violin on the right side
  geom_half_point(side = "l", alpha = 0.1, size = 2, color = "lightblue") +      # Points (the "rain") on the left side
  geom_half_boxplot(side = "l", width = 0.2, alpha = 0.8, outlier.shape = NA) +  # Half-boxplot on the left side
  coord_flip() +  # Flip coordinates so that scenarios appear on the y-axis
  scale_x_discrete(limits = c("Status Quo", "Housing Intervention")) +
  scale_y_continuous(labels = comma, limits = c(325000, 425000), breaks = seq(325000, 425000, by = 25000)) +
  labs(
    x = "Scenario",
    y = "Total Cost (2024 USD)") +
  theme_minimal(base_size = 14) +
  theme(
    # plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
    axis.title = element_text(size = 10, face = "bold"),
    axis.text = element_text(size = 10),
    legend.title = element_text(size = 10, face = "bold"),
    legend.text = element_text(size = 10),
    strip.text = element_text(size = 12, face = "bold"),
    panel.background = element_rect(fill = "transparent", color = NA),
    plot.background  = element_rect(fill = "transparent", color = NA),
    legend.position  = "top"
    #panel.grid.major = element_line(color = "grey80"),
    #panel.grid.minor = element_line(color = "grey90")
    #panel.grid.minor = element_blank()
  )

# Display the Total Cost plot
print(p_cost)
```

### Total QALYs

```{r}
df_long_qaly <- melt(
  dt_iter, 
  id.vars = "iteration", 
  measure.vars = c("total_QALY_base", "total_QALY_alt"),
  variable.name = "scenario", 
  value.name = "total_QALY"
)

df_long_qaly[, scenario := factor(
  scenario, 
  levels = c("total_QALY_base", "total_QALY_alt"),
  labels = c("Status Quo", "Housing Intervention")
)]

# Create a half‐violin (raincloud) plot for total QALYs
p_qaly <- ggplot(df_long_qaly, aes(x = scenario, y = total_QALY, fill = scenario)) +
  geom_half_violin(side = "r", alpha = 0.6, trim = TRUE, color = NA) +
  geom_half_point(side = "l", alpha = 0.1, size = 2, color = "lightblue") +
  geom_half_boxplot(side = "l", width = 0.2, alpha = 0.8, outlier.shape = NA) +
  coord_flip() +
  scale_x_discrete(limits = c("Status Quo", "Housing Intervention")) +
  scale_y_continuous(limits = c(3.5, 5), breaks = seq(3.5, 5, by = 0.25)) +
  labs(
    x = "Scenario",
    y = "Total QALYs") +
  theme_minimal(base_size = 14) +
  theme(
    #plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
    axis.title = element_text(size = 10, face = "bold"),
    axis.text = element_text(size = 10),
    legend.title = element_text(size = 10, face = "bold"),
    legend.text = element_text(size = 10),
    strip.text = element_text(size = 12, face = "bold"),
    panel.background = element_rect(fill = "transparent", color = NA),
    plot.background  = element_rect(fill = "transparent", color = NA),
    legend.position  = "top"
    #panel.grid.major = element_line(color = "grey80"),
    #panel.grid.minor = element_line(color = "grey90")
    #panel.grid.minor = element_blank()
  )

print(p_qaly)
```
